Surface-driven registration method for the structure-informed segmentation of diffusion MR images\tnoteref{t1}

### 2016-03-02 Publications/2015-NeuroImage/responseLetter3.tex

\usepackage{xcite}
\externalcitedocument{00-main}

\usepackage[colorlinks=true]{hyperref}
\usepackage{doi}
\usepackage{nameref}

\usepackage{amsmath}
\usepackage{amssymb}

\newcounter{reviewpoint}
\makeatletter
\newenvironment{reviewintro}%
{\begingroup%
 \color{black!60}
 \fontshape{it}\selectfont %
}
{\endgroup\par\medskip}

\newenvironment{reviewpoint}%
{\refstepcounter{reviewpoint}\par\medskip\vspace{3ex}\hrule\vspace{1.5ex}\par\noindent%
   {\fontseries{b}\selectfont Comment \arabic{reviewpoint}:}
   \begingroup%
   \color{black!60}
   \fontshape{it}\selectfont %

@@

\newcommand{\reply}{\par\fontshape{n}\selectfont \noindent \textbf{Reply}:\ }

\setlrmarginsandblock{3cm}{2.5cm}{*}
\setulmarginsandblock{2.5cm}{2.5cm}{*}
\checkandfixthelayout

\begin{document}
\hypersetup{linkcolor=black!60, citecolor=black!60, urlcolor=black!60}

\section*{Response letter to the manuscript no. NIMG-15-2877R1}

\bigskip
\noindent We thank the editors and the reviewers for revising our manuscript no. NIMG-15-2877R1 submitted to NeuroImage.
We appreciate the comments made by both reviewers, and considered them fully to improve our paper.
Please, find in the following paragraphs our point-by-point answer to the concerns raised by the reviewers,
  along with the description of the corresponding changes in the manuscript.
Text modifications in the manuscript are highlighted with a new font color, and attached notes indicate 
  which reviewer and comment the corresponding change is addressing to.

\bigskip
\bigskip
\subsection*{Reviewer \#1:}
\begin{reviewintro}
In general, I think the authors have done a good job of restructuring the paper to qualify the goals and implementation of the method. I have a few points.
\end{reviewintro}
\begin{reviewpoint}
The authors suggest that diffusion tractography is typically performed by mapping segmentations to the data. This is becoming less true. The HCP for one performs whole brain connectivity analysis. Nevertheless, these frameworks still require registration to structural space for projection to surface mesh representations and/or coaligning data. This may be worth mentioning or clarifying.
\end{reviewpoint}
\reply{%
We have introduced this discussion in lines 351--355.
}
\begin{reviewpoint}
I still find figure 2 really difficult to read anything from. It's not easy to read the labels in the coloured text, nor see the different histograms in figure 2 A. This is partly because many of the colours are light very similar i.e. for WM and bst. It does occur to me that not all distributions change that greatly with registration. Maybe stripping back figure A to only include WM, GM and possibly VdGM might help?
\end{reviewpoint}
\reply{%
We have made editions in this figure to make it more clear.
}
\begin{reviewpoint}
I find the clarification of how tratography might be unwarped in the response very helpful, but some clarification of the manuscript might be helpful. What do the authors mean by ``the resampling of the DWIs introduced by the correction method can be avoided, either by performing the posterior processing on the native dMRI space''. How does the native dMRI space differ from the DWIs here, or what is meant by posterior processing.
\end{reviewpoint}
\reply{%
We have reformulated that particular sentence (lines 57--59) and added a corresponding discussion in lines 348--356.
These editions are highly related to Comments \ref{com:1}, \ref{com:5} and \ref{com:6}.
}
\begin{reviewpoint}
In the appendix I think the B-spline variable (Eqs A.7 A.8)  should be defined in the text again
\end{reviewpoint}
\reply{%
We have defined introduced the definition of $\mathbf{g}_k$ in lines 122--123, right before the
  reference to the mentioned equations.
}

% ------------------------------------------------------------------- Reviewer 3
\bigskip
\bigskip
\subsection*{Reviewer \#3:}

\begin{reviewintro}
This manuscript is much improved, in my opinion. However, there are some high-level clarifications that are present in the author response letter, but that are not clear to the reader in the manuscript. I also noticed a few small things to fix. My suggestions follow.
\end{reviewintro}

\begin{reviewpoint}
Please clarify that the authors do not consider this to be an EPI distortion correction method, but rather a surface alignment method or similar. This was not clear to me from reading the paper the first time. I assumed EPI distortion correction was a primary goal, but it seems rather that the primary goal is to accurately segment the distorted diffusion images. In fact, the distortions are not corrected in the EPI data.
\end{reviewpoint}
\reply{%
We have explicitly stated this in a new sentence (lines 348--349).
}
\begin{reviewpoint}
Please discuss the fact that when this method is applied, the tractography is done using data that have not been corrected for EPI distortions.
\end{reviewpoint}
\reply{%
Right after the previous edition, in lines 349--350 we introduce the proposal of this methodological variation.
The discussion and proposal of a future research line is then done in lines 355--356, after some further details added
  to address the \autoref{com:1}.
}
\begin{reviewpoint}
For the reader, the overall comparison between the two methods is confusing in the introduction, where it is proposed to register surfaces to diffusion image space, yet the comparison is b0 to T2. A more appropriate way to present this comparison is t2 to b0, so that the result is also in the diffusion image space, if this is what was done. This is in fact a common processing step, so I assume this is how the experiments were performed. Please clarify.
\end{reviewpoint}
\reply{%
}
\begin{reviewpoint}
In Figure 6 A, is this output coordinate system the diffusion coordinate system as expected from the description of the method? So then the output in T2B is T2 to b0?
\end{reviewpoint}
\reply{%
}
\begin{reviewpoint}
The ``undistorted'' HCP data is not fully defined in the ``Real datasets'' section. Because this undistortion method serves as the ground truth for the presented experiments, it should be explained briefly how the undistortion was done for the HCP data. One sentence is sufficient, and will allow the reader better understand the results of this paper.
\end{reviewpoint}
\reply{%
We have introduced this description in lines 183--187, along with including the detail of the actual HCP release of the data used in l. 177.
}
\begin{reviewpoint}
Minor points:

The graphical abstract says subpixel accuracy. I think you mean subvoxel (there are no pixels really in 3D MRI images).

``preliminar DTI'' -> preliminary
\end{reviewpoint}
\reply{%
We have updated the graphical abstract with the term ``subvoxel''.
We have replaced the six appearances of the term ``pixel'' by ``voxel'', including the ``subpixel'' present in line 61.
We also have fixed the misspelling of ``preliminar'' of line 339.
}

### 2016-03-02 Publications/2015-NeuroImage/03-introduction.tex
getting submission ready

Here, we present a segmentation and surface-to-volume registration method
  called \regseg{}, and show its usefulness in mapping anatomical information from structural
  space into native \gls*{dmri} space to aid subsequent processing steps
  (reconstruction, tractography and network building using a cortical parcellation).
The underlying hypothesis is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
A specific set of reference surfaces extracted from the same subject initialize
  the 3D active contours, which evolve searching for homogeneous regions in the multivariate
  target-image.
We apply \regseg{} to segment \gls*{dmri} data by mapping
  a set of nested surfaces extracted from a structural image (e.g. \gls*{t1}) to
  a bivariate target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  deals with the nonlinear warping induced by \gls*{epi} distortions.
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol in order to estimate the deformation field;
  2) depending on the application, the resampling of the \glspl*{dwi}
    introduced by the correction method can be avoided, either by performing the posterior
    processing on the native \gls*{dmri} space or applying the unwarping on the tractography
    itself; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the sub-pixel accuracy
  in registration.
Then, we evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to
  the corresponding \gls*{t2} image of the same subject.
This approach is the first step of the \gls*{t2b} correction
  methods \citep{kybic_unwarping_2000}.
We reproduce the settings and implementation of a widely used diffusion
  processing software \citep[\emph{ExploreDTI},][]{leemans_exploredti_2009}.
With this comparison, we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

>>>>

Here, we present a segmentation and surface-to-volume registration method
  called \regseg{}, and show its usefulness in mapping anatomical information from structural
  space into native \gls*{dmri} space to aid subsequent processing steps
  (reconstruction, tractography and network building using a cortical parcellation).
The underlying hypothesis is that the registration and segmentation problems
  in \gls*{dmri} can be solved simultaneously.
To implement \regseg{} we first establish an active-contours without edges
  \citep{chan_active_2001} segmentation framework.
A specific set of reference surfaces extracted from the same subject initialize
  the 3D active contours, which evolve searching for homogeneous regions in the multivariate
  target-image.
We apply \regseg{} to segment \gls*{dmri} data by mapping
  a set of nested surfaces extracted from a structural image (e.g. \gls*{t1}) to
  a bivariate target-volume comprehending the \gls*{fa} and \gls*{adc} maps.
The evolution of the surfaces is supported by a B-spline basis, optimized
  iteratively using a descent approach driven by shape-gradients
  \citep{besson_dream2s_2003,herbulot_segmentation_2006}.
Therefore, \regseg{} establishes a registration framework that actually
  deals with the nonlinear warping induced by \gls*{epi} distortions.
\Regseg{} integrates the benefits of segmentation and registration methods together and
  exploits the multivariate nature of \gls*{dmri} data to contribute in the proposed
  application on three key aspects:
  1) the surfaces are typically extracted from the \gls*{t1} image of the same subject,
    therefore \regseg{} does not require additional MR acquisitions to the minimal
    \gls*{dmri} protocol in order to estimate the deformation field;
  2) \revcomment[R\#1-C2]{%
    alternatively to the typical design of the processing flow, the information from the
    reference \gls*{t1} can precisely be mapped onto the distorted \gls*{dmri} space,
    avoiding the interpolation of the \glspl*{dwi} required by unwarping the diffusion data; and} 
%    depending on the application, the resampling of the \glspl*{dwi}
%    introduced by the correction method can be avoided, either by performing the posterior
%    processing on the native \gls*{dmri} space or applying the unwarping on the tractography
%    itself; and
  3) \regseg{} increases the geometrical accuracy of the overall process.
In this paper, we first verify the functionality of the method and the \regseg{}
  implementation using a set of digital phantoms, demonstrating the
  \revcomment[R\#3-C10]{subvoxel} accuracy in registration.
Then, we evaluate \regseg{} on real \gls*{dmri} datasets, using a derivation of our
  instrumentation framework \citep{esteban_simulationbased_2014} which simulates
  known and realistic \gls*{epi} distortions.
We also compare \regseg{} and a nonlinear registration method to map the \lowb{} to
  the corresponding \gls*{t2} image of the same subject.
This approach is the first step of the \gls*{t2b} correction
  methods \citep{kybic_unwarping_2000}.
We reproduce the settings and implementation of a widely used diffusion
  processing software \citep[\emph{ExploreDTI},][]{leemans_exploredti_2009}.
With this comparison, we demonstrate how \regseg{} achieves higher accuracy with the
  simultaneous registration and segmentation process.

### 2016-03-02 Publications/2015-NeuroImage/04-methods.tex
getting submission ready

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:regseg-bayes_rule}
  \end{equation}
  where $P(M \mid U,\omegaset)$ is the data likelihood.
Since $\omegaset$ are mapped by $U$, we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.
The best estimate $\hat{U}$ then satisfies the maximum a posteriori criterion
  and it aligns $\gammaset_R$ into $M$.
First, we assume independence between pixels, and thus we break down the
  global data likelihood into a product of pixel-wise conditional probabilities:

@@

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \big[ \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r}' \big] \,
  + \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}'
  \Big\},
  \label{eq:regseg-gradient_descent}
  \end{align}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$,
  and $\{\boldsymbol{\alpha}, \boldsymbol{\beta}\}$ are the expected variances along
  the imaging axes of the displacement field and its gradient, respectively.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The derivative of the data term is computed using explicit shape gradients (see \ref{app:shape_gradients}),
  which finally lead to obtain vertex-wise speeds of the gradient $\bar{s}_i$ as
  illustrated in \autoref{fig:regseg-method}.
Using the expressions \eqref{eq:regseg-gradient_wshape} and 
  \eqref{eq:regseg-shape_gradient_final} given in \ref{app:shape_gradients}
  and introducing the analytical derivative of the regularization term, then
  the \autoref{eq:regseg-gradient_descent} is reformulated as:

@@

\paragraph*{Data}
To evaluate \regseg{} using real \gls*{dmri} data obtained from human brains,
  we collected 16 subjects from the \gls*{hcp} database.
The original acquisitions are released within ``unprocessed'' packages, whereas
  the ``minimally preprocessed'' packages contain the corresponding images after
  some processing (correction for several artifacts, brain-extraction, spatial
  normalization, etc.).
We refer the reader to \citet{essen_human_2012} for exact details of the acquisition
  parameters and \citet{glasser_minimal_2013} for the preprocessing issues.
These datasets comprise a large set of images, including \gls*{t1}, \gls*{t2}, and
  multi-shell \gls*{dmri} images.
Since we obtained the \gls*{dmri} data from the minimally preprocessed package, these
  images are corrected for \gls*{epi} distortions and spatially normalized in
  \gls*{t1} space.

@@

  \begin{equation}
  u_\text{PE} = \frac{\gamma \, T_{acq}\, s_\text{PE}}{2\pi}\Delta B_0(\vec{r})\text{ [mm]},
  \label{eq:regseg-fieldmap}
  \end{equation}
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time, and
  $s_\text{PE}$ is the pixel size along \gls*{pe}.
Certain \acrshort*{mri} sequences are designed to estimate $\Delta B_0$, thereby obtaining
  the so-called fieldmap.
We derived the deformation $U_\text{true}$ from the fieldmap image released with
  the corresponding packages of each dataset in the \gls*{hcp}.
The fieldmap was unwrapped\footnote{fieldmaps are phase maps, which are intrinsically clipped in the interval
  of $[-\pi, \pi)$ [rads] or [rads/s].} and smoothed before applying \eqref{eq:regseg-fieldmap}.
Next, the original \gls*{dmri} was warped using the resulting displacement field and fed into
  a pipeline to process the corresponding \gls*{dti}, thereby computing the derived scalars of
  interest (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.

@@

\paragraph*{Cross-comparison} %
A dual workflow to the general evaluation used for \regseg{} (\autoref{fig:regseg-evworkflows}),
  was employed to integrate the alternate \gls*{t2b} registration scheme.
We reproduced the solution and settings provided with \emph{ExploreDTI}
  \citep{leemans_exploredti_2009}, which is a widely used toolkit for tractography analysis of
  \gls*{dti}.
\emph{ExploreDTI} internally employs \emph{elastix} \citep{klein_elastix_2010} to
  perform registration.
The deformation field is correspondingly restricted to the \gls*{pe} direction.
The settings file for \emph{elastix} is also
  available\footnote{\url{https://github.com/oesteban/RegSeg/blob/master/Scripts/pyacwereg/data/t2b_elastix_y.txt}}.

>>>>

  \begin{equation}
  P(U \mid M, \omegaset ) = \frac{P(M \mid U,\omegaset )\, P(U)}{P(M)},
  \label{eq:regseg-bayes_rule}
  \end{equation}
  where $P(M \mid U,\omegaset)$ is the data likelihood.
Since $\omegaset$ are mapped by $U$, we simplify
  $P(U, \omegaset) = P(U) \implies P(M \mid U,\omegaset) = P(M \mid U)$.
The best estimate $\hat{U}$ then satisfies the maximum a posteriori criterion
  and it aligns $\gammaset_R$ into $M$.
First, we assume independence between \revcomment[R\#3-C10]{voxel}s, and thus we break down the
  global data likelihood into a product of \revcomment[R\#3-C10]{voxel}-wise conditional probabilities:

@@

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \big[ \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r}' \big] \,
  + \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}'
  \Big\},
  \label{eq:regseg-gradient_descent}
  \end{align}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$,
  and $\{\boldsymbol{\alpha}, \boldsymbol{\beta}\}$ are the expected variances along
  the imaging axes of the displacement field and its gradient, respectively.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The derivative of the data term is computed using explicit shape gradients (see \ref{app:shape_gradients}),
  which finally lead to obtain vertex-wise speeds of the gradient $\bar{s}_i$ as
  illustrated in \autoref{fig:regseg-method}.
\revcomment[R\#1-C4]{%
The shape gradient contributions $\vec{g}_k$ on the field coefficients $\vec{u}_k$
  can then be computed }%
using the expressions \eqref{eq:regseg-gradient_wshape} and 
  \eqref{eq:regseg-shape_gradient_final} given in \ref{app:shape_gradients}.
Finally, introducing the analytical derivative of the regularization term,
  the \autoref{eq:regseg-gradient_descent} is reformulated as:

@@

\paragraph*{Data}
To evaluate \regseg{} using real \gls*{dmri} data obtained from human brains,
  we collected 16 subjects from the \revcomment[R\#3-C9]{Q3 Release of the}
  \gls*{hcp} database.
The original acquisitions are released within ``unprocessed'' packages, whereas
  the ``minimally preprocessed'' packages contain the corresponding images after
  some processing (correction for several artifacts, brain-extraction, spatial
  normalization, etc.).
We refer the reader to \citet{essen_human_2012} for exact details of the acquisition
  parameters and \citet{glasser_minimal_2013} for the preprocessing issues.
These datasets comprise a large set of images, including \gls*{t1}, \gls*{t2}, and
  multi-shell \gls*{dmri} images.
Since we obtained the \gls*{dmri} data from the minimally preprocessed package, these
  images are corrected for \gls*{epi} distortions and spatially normalized in
  \gls*{t1} space.
\revcomment[R\#3-C9]{In the Q3 Release of the \gls*{hcp}, the \gls*{dmri} session includes
  six runs, two runs for each of three different gradient tables, and each table is acquired
  once with right-to-left and left-to-right encoding polarities, respectively.
Then, the diffusion datasets with opposed polarities are corrected for susceptibility
  distortions using the \emph{TOPUP} tool \citep{andersson_how_2003}
  before producing the released ``minimally preprocessed'' data.}.

@@

  \begin{equation}
  u_\text{PE} = \frac{\gamma \, T_{acq}\, s_\text{PE}}{2\pi}\Delta B_0(\vec{r})\text{ [mm]},
  \label{eq:regseg-fieldmap}
  \end{equation}
where $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time, and
  $s_\text{PE}$ is the \revcomment[R\#3-C10]{voxel} size along \gls*{pe}.
Certain \acrshort*{mri} sequences are designed to estimate $\Delta B_0$, thereby obtaining
  the so-called fieldmap.
We derived the deformation $U_\text{true}$ from the fieldmap image released with
  the corresponding packages of each dataset in the \gls*{hcp}.
The fieldmap was unwrapped\footnote{fieldmaps are phase maps, which are intrinsically clipped in the interval
  of $[-\pi, \pi)$ [rads] or [rads/s].} and smoothed before applying \eqref{eq:regseg-fieldmap}.
Next, the original \gls*{dmri} was warped using the resulting displacement field and fed into
  a pipeline to process the corresponding \gls*{dti}, thereby computing the derived scalars of
  interest (\gls*{fa} and \gls*{adc}) using \emph{MRtrix} \citep{tournier_mrtrix_2012}.

@@

\paragraph*{Cross-comparison} %
A dual workflow to the general evaluation used for \regseg{} (\autoref{fig:regseg-evworkflows}),
  was employed to integrate the alternate \gls*{t2b} registration scheme.
We reproduced the solution and settings provided with \emph{ExploreDTI}
  \citep{leemans_exploredti_2009}, which is a widely used toolkit for tractography analysis of
  \gls*{dti}.
\emph{ExploreDTI} internally employs \emph{elastix} \citep{klein_elastix_2010} to
  perform registration.
The deformation field is correspondingly restricted to the \gls*{pe} direction.
The settings file for \emph{elastix} is also
  available\footnote{\url{https://github.com/oesteban/RegSeg/blob/master/Scripts/pyacwereg/data/t2b_elastix_y.txt}}.
\revcomment[R\#3-C7\\R\#3-C7]{%
In this registration scheme, the \gls*{t2} image is the \emph{reference} and
  the \lowb{} plays the role of \emph{moving} image.
Therefore, the transform is defined in the coordinate system of the \gls*{t2} image,
  and for each point in this space provides the location of the corresponding feature
  in the \lowb{} image.
Since the surfaces are defined in the \gls*{t2} --\emph{reference}-- space, their coordinates
  can be mapped to the \lowb{} space using this transform, obtaining the 
  \emph{distorted surfaces} corresponding to the \lowb{}-to-\gls*{t2} registration.}
